Using structural and evolutionary information to detect and correct pyrosequencing errors in non-coding RNAs

### 2014-02-04 JCB/results_JCB.tex
JCBFinal

\subsection{Implementation}
The software was implemented in Python2.7 using the \textit{mpmath}~\cite{mpmath} library
for  arbitrary floating point precision. The source code is freely available at:

@@

{\centering \url{http://csb.cs.mcgill.ca/RNApyro}\\}
The time benchmarks were performed on an AMD Opteron(tm) Processor 6278  at 2.4 GHz with cache of 512 KB.20 cores and 175 GB ram.
Since typical use-cases of \RNApyro require efficiency and scalability, we present in TableFigure~\ref{fig:time}
typical runtimes required to compute the probabilities for  every nucleotide at every positions for a vast set of parameters.
For those tests, both the multiple sequence alignment and the reference sequence were generated uniformly at random,
based on a realistic random secondary structure, generated as described in Levin~\emph{et al}~\cite{Levin:2012kx}.
%\TODOVlad{Add more extensive time benchmark, as a plot}
%\TODOVlad{Is time benchmark still on Mac mini or mcbbigram?}

\begin{figure}[t]
{\centering \includegraphics[width=.5\linewidth]{figures/TimeBenchmark}\\}

@@

We built our data set from the 5S rRNA multiple sequence alignment (MSA) available in the Rfam Database 11.0 (Rfam id: \texttt{RF00001}).
Since our software does not currently implement gaps (mainly because scoring indels is a challenging issue that cannot be fully addressed
in this work),  we clustered together the sequences with identical gap locations. From the $54$ MSAs without gap produced, we selected the
biggest MSA  which contains $130$ sequences (out of $712$ in the original Rfam MSA). Then, in order to avoid overfitting, we used \texttt{cd-hit}
\cite{Li:2006fk} to remove sequences with more than 80\% of sequence similarity. This operation resulted in a data set of $45$ sequences.

@@

\begin{figure}
\centering
	\includegraphics[width=\textwidth]{subfigs_perform.png}\\

@@

We gathered the seed sequences of the bacterial multiple sequence alignment (MSA) retrieved from the RFAM Database 11.0 (Rfam id: RF00177)~\cite{gardner2011rfam}. This alignment is composed of 93 sequences, whose length varies between 1461 and 1568 nucleotides, and has an average pairwise sequence identity of $69\%$. We used the pseudoknot-free version of the consensus secondary structure. A secondary structure for a specific reference sequence was obtained by simply mapping the structure back to sequence, i.e. by removing any base-pair having at least one partner involved in a gapped position. For similar reasons, we locally excluded from our calculation of the isostericity contribution for a given base-pair, described in Section~\ref{sec:model}, any sequence featuring at least a gap on the corresponding positions.

To simulate sequencing errors, we used the next-generation sequencing read simulator 
{\tt ART}~\cite{huang2012art}. The \emph{Illumina technology} setting was chosen as the main error mode, generating reads of $75~\text{bps}$, featuring mostly base substitution errors. Reads were mapped back to the original sequence, and a consensus sequence was determined, from the sequencing output, by a simple majority vote in the case of multiple coverage. Uncovered regions were simply generated at random. The average  rate of errors observed in the final consensus sequence was empirically estimated to represent $2.4\%$, $0.9\%$ and $0.6\%$ of the reference sequence, for prescribed sequencing coverages of $5$, $10$ and $15$ fold respectively.

@@

 \begin{figure}
\centering
\begin{subfigure}{.33\textwidth}
  \centering
  \includegraphics[width=\linewidth]{figures/5fold-a}
  \caption{$5-$fold coverage}
\end{subfigure}%
\begin{subfigure}{.33\textwidth}
  \centering
  \includegraphics[width=\linewidth]{figures/10fold-a}
    \caption{$10-$fold coverage}
\end{subfigure}
\begin{subfigure}{.33\textwidth}
  \centering
  \includegraphics[width=\linewidth]{figures/15fold-a}
    \caption{$15-$fold coverage}
\end{subfigure}
\caption{Performance of error-correction over all positions. Subfigures show accuracy when
ART fold parameter is set to 5, 10 and $15-$fold coverage. We also analyze the impact of the parameter $\alpha$ distributing the weights of stacking pair energies vs isostericity scores and use values ranging of $\alpha = \{0, 0.5, 0.8, 1.0\}$. The AUC is indicated in the legend of the figures. Each individual ROC curve represent the average performance over 30 experiments.}

@@

 \begin{figure}
\centering
\begin{subfigure}{.33\textwidth}
  \centering
  \includegraphics[width=\linewidth]{figures/5fold-b}
  \caption{$5-$fold coverage}
\end{subfigure}%
\begin{subfigure}{.33\textwidth}
  \centering
  \includegraphics[width=\linewidth]{figures/10fold-b}
    \caption{$10-$fold coverage}
\end{subfigure}
\begin{subfigure}{.33\textwidth}
  \centering
  \includegraphics[width=\linewidth]{figures/15fold-b}
    \caption{$15-$fold coverage}
\end{subfigure}
\caption{Performance of error-correction restricted to structured positions (i.e. involved in a canonical base pair). Subfigures show accuracy when
ART fold parameter is set to 5, 10 and $15-$fold coverage. We also analyze the impact of the parameter $\alpha$ distributing the weights of stacking pair energies vs isostericity scores and use values ranging of $\alpha = \{0, 0.5, 0.8, 1.0\}$. The AUC is indicated in the legend of the figures. Each individual ROC curve represent the average performance over at least 10 experiments.}
\label{fig:16s_bp}
\end{figure}

@@

An interesting feature is that best results are almost always when $\alpha=0$, i.e. all the contribution
comes from the isostericity.
% A notable exception is when the number of mutation is underestimated, 
%in that case best performances were observed for non-null values of $\alpha$.

>>>>

\subsection{Implementation}
The software was implemented in Python2.7 using the \textit{mpmath}~\citep{mpmath} library
for  arbitrary floating point precision. The source code is freely available at:

@@

{\centering \url{http://csb.cs.mcgill.ca/RNApyro}\\}
The time benchmarks were performed on an AMD Opteron(tm) Processor 6278  at 2.4 GHz with cache of 512 KB.20 cores and 175 GB ram.
Since typical use-cases of \RNApyro require efficiency and scalability, we present in TableFigure~\ref{fig:time}
typical runtimes required to compute the probabilities for  every nucleotide at every positions for a vast set of parameters.
For those tests, both the multiple sequence alignment and the reference sequence were generated uniformly at random,
based on a realistic random secondary structure, generated as described in~\cite{Levin:2012kx}.
%\TODOVlad{Add more extensive time benchmark, as a plot}
%\TODOVlad{Is time benchmark still on Mac mini or mcbbigram?}

\begin{figure}[t]
{\centering \includegraphics[width=.5\linewidth]{Figure4-TimeBenchmark}\\}

@@

We built our data set from the 5S rRNA multiple sequence alignment (MSA) available in the Rfam Database 11.0 (Rfam id: \texttt{RF00001}).
Since our software does not currently implement gaps (mainly because scoring indels is a challenging issue that cannot be fully addressed
in this work),  we clustered together the sequences with identical gap locations. From the $54$ MSAs without gap produced, we selected the
biggest MSA  which contains $130$ sequences (out of $712$ in the original Rfam MSA). Then, in order to avoid overfitting, we used \texttt{cd-hit}
\citep{Li:2006fk} to remove sequences with more than 80\% of sequence similarity. This operation resulted in a data set of $45$ sequences.

@@

\begin{figure}
\centering
	\includegraphics[width=\textwidth]{Figure5-subfigs_perform}\\

@@

We gathered the seed sequences of the bacterial multiple sequence alignment (MSA) retrieved from the RFAM Database 11.0 (Rfam id: RF00177)~\citep{gardner2011rfam}. This alignment is composed of 93 sequences, whose length varies between 1461 and 1568 nucleotides, and has an average pairwise sequence identity of $69\%$. We used the pseudoknot-free version of the consensus secondary structure. A secondary structure for a specific reference sequence was obtained by simply mapping the structure back to sequence, i.e. by removing any base-pair having at least one partner involved in a gapped position. For similar reasons, we locally excluded from our calculation of the isostericity contribution for a given base-pair, described in Section~\ref{sec:model}, any sequence featuring at least a gap on the corresponding positions.

To simulate sequencing errors, we used the next-generation sequencing read simulator 
{\tt ART}~\citep{huang2012art}. The \emph{Illumina technology} setting was chosen as the main error mode, generating reads of $75~\text{bps}$, featuring mostly base substitution errors. Reads were mapped back to the original sequence, and a consensus sequence was determined, from the sequencing output, by a simple majority vote in the case of multiple coverage. Uncovered regions were simply generated at random. The average  rate of errors observed in the final consensus sequence was empirically estimated to represent $2.4\%$, $0.9\%$ and $0.6\%$ of the reference sequence, for prescribed sequencing coverages of $5$, $10$ and $15$ fold respectively.

@@

 \begin{figure}[t]
\centering
\begin{subfigure}{.33\textwidth}
  \centering
  \includegraphics[width=\linewidth]{Figure6-5fold-a}
  \caption{$5-$fold coverage}
\end{subfigure}%
\begin{subfigure}{.33\textwidth}
  \centering
  \includegraphics[width=\linewidth]{Figure7-10fold-a}
    \caption{$10-$fold coverage}
\end{subfigure}
\begin{subfigure}{.33\textwidth}
  \centering
  \includegraphics[width=\linewidth]{Figure8-15fold-a}
    \caption{$15-$fold coverage}
\end{subfigure}
\caption{Performance of error-correction over all positions. Subfigures show accuracy when
ART fold parameter is set to 5, 10 and $15-$fold coverage. We also analyze the impact of the parameter $\alpha$ distributing the weights of stacking pair energies vs isostericity scores and use values ranging of $\alpha = \{0, 0.5, 0.8, 1.0\}$. The AUC is indicated in the legend of the figures. Each individual ROC curve represent the average performance over 30 experiments.}

@@

 \begin{figure}
\centering
\begin{subfigure}{.33\textwidth}
  \centering
  \includegraphics[width=\linewidth]{Figure9-5fold-b}
  \caption{$5-$fold coverage}
\end{subfigure}%
\begin{subfigure}{.33\textwidth}
  \centering
  \includegraphics[width=\linewidth]{Figure10-10fold-b}
    \caption{$10-$fold coverage}
\end{subfigure}
\begin{subfigure}{.33\textwidth}
  \centering
  \includegraphics[width=\linewidth]{Figure11-15fold-b}
    \caption{$15-$fold coverage}
\end{subfigure}
\caption{Performance of error-correction restricted to structured positions (i.e. involved in a canonical base pair). Subfigures show accuracy when
ART fold parameter is set to 5, 10 and $15-$fold coverage. We also analyze the impact of the parameter $\alpha$ distributing the weights of stacking pair energies vs isostericity scores and use values ranging of $\alpha = \{0, 0.5, 0.8, 1.0\}$. The AUC is indicated in the legend of the figures. Each individual ROC curve represent the average performance over at least 10 experiments.}
\label{fig:16s_bp}
\end{figure}

@@

An interesting feature is that best results are almost always obtained when $\alpha=0$, i.e. when all the contribution
comes from the isostericity.
% A notable exception is when the number of mutation is underestimated, 
%in that case best performances were observed for non-null values of $\alpha$.
